---
output:
   md_document:
    fig_width: 8
    fig_height: 3
---

```{r initialisation, echo = TRUE, message = F, warning = F}
library(ggthemes)
library(ggplot2)
library(data.table)
library(ggTimeSeries)
```

## Line Charts Legacy
IoT devices generate a lot of sequential data over time, also called time series data. Legacy portrayals of such data would centre around line charts. Line charts have reportedly been around since the early 1700s (source: Wikipedia) and we have nothing against them.  They facilitate trend detection and comparison, are simple to draw, and easy to understand; all in all a very well behaved visualisation. Their use is widespread from the heartbeat monitor in a hospital to the multiple-monitor display at a bank.

```{r excel97_line, ext = 'png', fig.align = 'center', echo = FALSE, message = F, warning = F}
set.seed(10)
df = data.frame(
   Time = 1:100,
   Signal = abs(
      c(
         cumsum(rnorm(100, 0, 3)), 
         cumsum(rnorm(100, 0, 4)), 
         cumsum(rnorm(100, 0, 1)), 
         cumsum(rnorm(100, 0, 2))
      )
   ),
   Variable = c(rep('a', 100), rep('b', 100), rep('c', 100), rep('d', 100)),
   VariableLabel = c(rep('Class A', 100), rep('Class B', 100), rep('Class C', 100), rep('Class D', 100))
)

Excel97Plot = ggplot(df, aes(x = Time, y = Signal, color = VariableLabel)) +
   geom_line() +
   geom_point() +
   theme_excel() + 
   scale_colour_excel()

print("Excel 97 look recreated in R with the ggthemes package")
plot(Excel97Plot)

```

However there are cases when the data scientist becomes more demanding and specific. Five alternatives available to such a data scientist are listed below. All of these options are available as `geom`s or packaged functions in the `ggplot2` based `ggTimeSeries` package.

Before that, setting a minimal theme -
```{r minimalTheme}
minimalTheme = theme_set(theme_bw(12))
minimalTheme = theme_update(
   axis.ticks = element_blank(), 
   legend.position = 'none',
   strip.background = element_blank(), 
   panel.border = element_blank(), 
   panel.background = element_blank(), 
   panel.grid = element_blank(), 
   panel.border = element_blank()
)

```

## Calendar heatmap
Available as `stat_calendar_heatmap` and `ggplot_calendar_heatmap`.

A calendar heatmap is a great way to visualise daily data. Its structure makes it easy to detect weekly, monthly, or seasonal patterns. Can you spot how the weekends seem to be different from the weekdays and how summer seems to be different from winter?

```{r calendar_heatmap, fig.align = 'center', echo = TRUE, message = F, warning = F}

# creating some data
set.seed(1)
dtData = data.table(
      DateCol = seq(
         as.Date("1/01/2014", "%d/%m/%Y"),
         as.Date("31/12/2015", "%d/%m/%Y"),
         "days"
      ),
      ValueCol = runif(730)
   )
dtData[, ValueCol := ValueCol + (strftime(DateCol,"%u") %in% c(6,7) * runif(1) * 0.75), .I]
dtData[, ValueCol := ValueCol + (abs(as.numeric(strftime(DateCol,"%m")) - 6.5)) * runif(1) * 0.75, .I]
dtData[, CategCol := letters[round(ValueCol * 7)]]

# base plot
p1 = ggplot_calendar_heatmap(
   dtData,
   'DateCol',
   'ValueCol'
)

# adding some formatting
p1 + 
   xlab('') + 
   ylab('') + 
   scale_fill_continuous(low = 'green', high = 'red') + 
   facet_wrap(~Year, ncol = 1)

# base plot
p2 = ggplot_calendar_heatmap(
   dtData,
   'DateCol',
   'CategCol'
)

# adding some formatting
p2 + 
   xlab('') + 
   ylab('') + 
   facet_wrap(~Year, ncol = 1)


```


## Horizon plots
Available as `stat_horizon` and `ggplot_horizon`.

Imagine an area chart which has been chopped into multiple chunks of equal height. If you overlay these chunks one on top of the the other, and colour them to indicate which chunk it is, you get a horizon plot. Horizon plots are useful when visualising y values spanning a vast range and / or trying to highlight outliers without losing context of variation in the rest of the data. 


```{r horizon, fig.align = 'center', echo = TRUE, message = F, warning = F}

# creating some data
set.seed(1)
dfData = data.frame(x = 1:1000, y = cumsum(rnorm(1000)))

# base plot
p1 = ggplot_horizon(dfData, 'x', 'y')

# adding some formatting
p1 + 
   xlab('') + 
   ylab('') + 
   scale_fill_continuous(low = 'green', high = 'red') + 
   coord_fixed( 0.5 * diff(range(dfData$x)) / diff(range(dfData$y)))
```


## Steamgraphs

Available as `stat_steamgraph`.

A steamgraph is a more aesthetically appealing version of a stacked area chart. It tries to highlight the changes in the data by placing the groups with the most variance on the edges, and the groups with the least variance towards the centre. This feature in conjunction with the centred alignment of each of the contributing areas makes it easier for the viewer to compare the contribution of any of the components across time.

```{r steamgraph, fig.align = 'center', echo = TRUE, message = F, warning = F}
# creating some data
set.seed(10)
df = data.frame(
   Time = 1:1000,
   Signal = abs(
      c(
         cumsum(rnorm(1000, 0, 3)), 
         cumsum(rnorm(1000, 0, 4)), 
         cumsum(rnorm(1000, 0, 1)),
         cumsum(rnorm(1000, 0, 2))
      )
   ),
   VariableLabel = c(rep('Class A', 1000), rep('Class B', 1000), rep('Class C', 1000), rep('Class D', 1000))
)

# base plot
p1 = ggplot(df, aes(x = Time, y = Signal, group = VariableLabel, fill = VariableLabel)) +
  stat_steamgraph()


# adding some formatting
p1 + 
   xlab('') + 
   ylab('') + 
   coord_fixed( 0.2 * diff(range(df$Time)) / diff(range(df$Signal)))

```


## Waterfall
Available as `stat_waterfall` and `ggplot_waterfall`.

A waterfall plot emphasis the changes in the time series.


```{r waterfall, fig.align = 'center', echo = TRUE, message = F, warning = F}
# creating some data
set.seed(1)
dfData = data.frame(x = 1:100, y = cumsum(rnorm(100)))

# base plot
p1 = ggplot_waterfall(
   dtData = dfData,
   'x',
   'y'
)

# adding some formatting
p1 + 
   xlab('') + 
   ylab('')
```


## Occurrence dot plot
Available as `stat_occurrence`.

This one is a favourite in infographics. For rare events, the reader will probably find it convenient to have the count of events encoded in the chart itself and this is the purpose served by the occurrence dot plot.


```{r occurrence_dotplot, fig.align = 'center', echo = TRUE, message = F, warning = F}
# creating some data
set.seed(1)
dfData = data.table(x = 1:100, y = floor(4 * abs(rnorm(100, 0 , 0.4))))

# base plot
p1 = ggplot(dfData, aes(x =x, y = y) )+
   stat_occurrence()

# adding some formatting   
p1 +
   xlab('') + 
   ylab('') + 
   coord_fixed(ylim = c(0,1 + max(dfData$y)))
```
